Think deep for reef resilience

### 2025-09-10 paper/paper_draft.qmd
Add cots control to discussion

Current reef restoration and intervention investigation efforts on the GBR, such as
deployment of coral seeding and fragment devices, target reefs with suitable sites in the
1-10m depth range [@howlett2021; @randall2021; @whitman2024]. While there are no explicit
justifications for this depth range, intervention depth is likely constrained by snorkelling
and diving deployment and monitoring logistical accessibility, visual identification of
sites from the surface, in combination with tourism return value. Our results suggest that
of the assessed reefs occurring within the 1-10m depth range (1736 reefs), 57% (991) of
reefs fall into the low and medium clusters (across all GCMs at the GBRMPA management region
scale), with many unable to maintain positive carbonate budgets under future conditions.
Partial dependence analysis revealed a range of depths from 10-15m where the probability of
a reef being clustered into the High cover cluster increased substantially. Additionally,
two-way partial dependence analysis revealed that high incoming connectivity is important
for High cluster assignment for reefs at depths of \>10m. Considering this 10-15m depth
range, and targetting well connected reefs, when assessing possible intervention sites may
reduce the exposure of planted and seeded corals to high thermal stress, while still
maintaining effective logistical access to sites. Managers are also likely to consider the
positive potential for flow on effects due to outgoing connectivity when selecting
intervention sites [@hock2017]. Within the context of reef depth, reefs with high depths and
high outgoing connectivity levels had the greatest probability of being clustered into high
cover clusters (@sfig-ts-cluster-pdp-twoway-outgoing). These results suggests that
interventions targetting deep reefs will also benefit from high incoming and outgoing
connectivity levels, allowing for high potential for site resilience and outgoing flow of
benefits and recovery. Transient, and difficult to measure, properties such as reef
connectivity are likely to increase uncertainty in intervention planning to a greater degree
than environmental characteristics such as reef depth. Overall, reef intervention planning
is a complex and challenging decision-making process [@anthony2020], and our results suggest
that intervention site depth may be an important factor to additionally consider when
assessing the long-term resilience of sites.

>>>>

Current reef restoration and intervention efforts on the GBR, such as deployment of coral
seeding, fragment devices and Crown of Thorns Starfish control, target reefs with suitable
sites in the 1-10m depth range [@howlett2021; @randall2021; @whitman2024]. While there are
no explicit justifications for this depth range, intervention depth is likely constrained by
snorkelling and diving deployment and monitoring logistical accessibility, visual
identification of sites from the surface, in combination with tourism return value. Our
results suggest that of the assessed reefs occurring within the 1-10m depth range (1736
reefs), 57% (991) of reefs fall into the low and medium clusters (across all GCMs at the
GBRMPA management region scale), with many unable to maintain positive carbonate budgets
under future conditions. Partial dependence analysis revealed a range of depths from 10-15m
where the probability of a reef being clustered into the High cover cluster increased
substantially. Additionally, two-way partial dependence analysis revealed that high incoming
connectivity is important for High cluster assignment for reefs at depths of \>10m.
Considering this 10-15m depth range, and targetting well connected reefs, when assessing
possible intervention sites may reduce the exposure of planted and seeded corals to high
thermal stress, while still maintaining effective logistical access to sites. Managers are
also likely to consider the positive potential for flow on effects due to outgoing
connectivity when selecting intervention sites [@hock2017]. Within the context of reef
depth, reefs with high depths and high outgoing connectivity levels had the greatest
probability of being clustered into high cover clusters
(@sfig-ts-cluster-pdp-twoway-outgoing). These results suggests that interventions targetting
deep reefs will also benefit from high incoming and outgoing connectivity levels, allowing
for high potential for site resilience and outgoing flow of benefits and recovery.
Transient, and difficult to measure, properties such as reef connectivity are likely to
increase uncertainty in intervention planning to a greater degree than environmental
characteristics such as reef depth. Overall, reef intervention planning is a complex and
challenging decision-making process [@anthony2020], and our results suggest that
intervention site depth may be an important factor to additionally consider when assessing
the long-term resilience of sites.
